---
title: "Bayesian Hierarchical Modelling for Mortality Mapping"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
---

## Introduction

This project is aimed to explore the use of hierarchical Bayesian models for mortality mapping. 

The goal is to estimate relative risk across geographical regions while accounting for variability and uncertainty in observed mortality counts.

I modeled observed counts using a Poisson likelihood with random effect for each area and performed inference using MCMC methods implemented in JAGS.

This approach allows for more stable estimates compared to classical methods such as Standard Mortality Ratios (SMR), particularly in regions with small sample sizes.



```{r,warning=FALSE,message=FALSE}
library(tidyverse)
library(RColorBrewer)
library(coda)
library(sf)
library(rjags)
library(lattice)
library(MCMCvis)
library(R2jags)



setwd("/Users/hassanainqazzaz/Downloads")

# Load the data

scot = read.csv("Scotland.csv")

surg= read.csv("surgical.csv")





```

```{r,warning=FALSE,message=FALSE}

ScotlandMap <- function(data, figtitle="",lower=NULL,upper=NULL){
  # local authority administrative areas
  names <- c("Skye-Lochalsh", "Banff-Buchan",  "Caithness",     "Berwickshire",  "Ross-Cromarty", "Orkney",        "Moray" ,       
  "Shetland",      "Lochaber",      "Gorden",        "Western Isles", "Sutherland",    "Nairn",         "Wigtown",      
  "NE Fife",       "Kincardine",    "Bedenoch",      "Ettrick",       "Inverness",     "Roxburgh",      "Angus",        
  "Aberdeen",      "Argyll-Bute",   "Clydesdale",    "Kirkcaldy" ,    "Dunfermline",   "Nithsdale",     "East Lothian", 
  "Perth-Kinross", "West Lothian",  "Cumnock-Doon",  "Stewartry",     "Midlothian",    "Stirling",      "Kyle-Carrick", 
  "Inverclyde",    "Cunninghame",   "Monklands",     "Dumbarton",     "Clydebank",     "Renfrew",       "Falkirk" ,     
  "Clackmannan",   "Motherwell",    "Edinburgh",     "Kilmarnock",    "East Kilbride" ,"Hamilton",      "Glasgow",      
  "Dundee" ,       "Cumbernauld",   "Bearsden" ,     "Eastwood" ,    "Strathkelvin",  "Tweeddale" ,    "Annandale") 
  
  # merging provided data with area names
  df <- data.frame(Name=names,Value = data)
  
  # Reading in Scotland shapefiles
  library(sf)
  Scotland <- st_read("/Users/hassanainqazzaz/Downloads/Scotland_data/Scotland.shp", quiet = TRUE)
  
  # Combining data and the shapefile
  Scotland_data <- merge(Scotland, df,
               by = 'Name')
  Scotland_data <- st_as_sf(Scotland_data)
  
  # Setting the colour limits
  if(is.null(lower)) lower=floor(min(data))
  if(is.null(upper)) upper=ceiling(max(data))
  
  # Creating the map
  ggplot(Scotland_data,
         aes(fill = Value)) +
    geom_sf(colour = NA) + theme_bw() +
    labs(x = 'Longitude',
         y = 'Latitude',
         fill = 'Value') +
    scale_fill_gradientn(colours = brewer.pal(9, 'RdPu'),
                         breaks = seq(lower,upper,length.out=5)) +
    ggtitle(label=figtitle)
}





```


## Exploratory Analysis: Standard Mortality Ratios (SMR)

As a baseline, I computed the Standard Mortality Ratio (SMR):

SMR = Observed / Expected

This provides a naive estimate of relative risk, but does not account for 
statistical uncertainty or variability across regions.

```{r}

# Calculate the Standard Mortality Ratios 
# (SMR=observed/expected) for each administrative area

scot$smr = scot$Observed/scot$Expected

# Plot the distribution

# Plot the distribution of the SMRs
ggplot(scot, aes(x = smr)) +
  geom_histogram(binwidth = 0.5,  fill = "red") +
  xlab("Standard Mortality Ratio") +
  ylab("Count") +
  ggtitle("Distribution of SMRs for lip cancer in Scotland")



ScotlandMap(scot$smr,
            figtitle="Scotland Standard mortality rate by area")




```



## Bayesian Hierarchical Model

To account for variability across regions, a hierarchical Bayesian model is implemented:

- Observed counts are assumed to follow a Poisson distribution
- The log-risk is modeled as a combination of a global intercept and area-specific random effects

This allows to produce more stable estimates.

Where:

- β₀ represents the overall log-relative risk
- θᵢ represents area-specific deviations (random effects)
- θᵢ ~ Normal(0, τ)

This hierarchical structure captures unobserved heterogeneity across regions.

```{r,warning=FALSE,message=FALSE}


# Specify model in JAGS
jags.mod <- function(){
  # Likelihood
  for (i in 1:n) {
    Obs[i] ~ dpois(mu[i])
    log(mu[i]) <- log(Exp[i]) + beta0 + theta[i]
    theta[i] ~ dnorm(0, tau)
    RR[i] <- exp(beta0) * exp(theta[i])
  }
  
  # Prior for beta0
  beta0 ~ dnorm(0, 1/1000)
  
  
 
  tau ~ dgamma(1, 0.05)
}



# Extract the data to be added to a list
n = nrow(scot)
Name = scot[,1]
Obs = scot[,2]
Exp = scot[,3]
smr = scot[,4]


# Convert data to list
jags.data = list("n" = n,  "Obs" = Obs, 
                 "Exp" = Exp)


# Parameters we want to monitor
jags.param <- c('theta' ,"beta0","tau","RR","mu")

# Specify initial values
inits1 <-  list('tau' = 20, 'beta0'= 0,theta=rep(0,n))
inits2 <- list('tau'=22,'beta0' = 0.5 ,theta=rep(0.5,n))
jags.inits <- list(inits1, inits2)


# Fit the JAGS model
jags.mod.fit = jags(data = jags.data , inits = jags.inits,
                     parameters.to.save = jags.param, 
                    n.chains =2,
                     n.iter = 10000,
                     n.burnin = 2500,n.thin=1, 
                    model.file = jags.mod, 
                    DIC=TRUE)


print(jags.mod.fit)

# Convert to mcmc
jags.mcmc = as.mcmc(jags.mod.fit)


## Gelman plot indicates that 1000-2500 should be burnin
gelman.plot(jags.mcmc)

# Gelman diagnostics indicate that convergence has been reached


# Summaries and traceplots
MCMCtrace(jags.mcmc,type = 'trace',ind = TRUE,pdf=F)
summary(jags.mcmc)

```

Using the gelman.plot function as visual aid the n.burnin was set at 2500 where the chains seemed to settle and n.iter at 10k. The chains of all paramters seem to have converged looking at the traceplots which show good mixing, also the gelam rubin diagnostic all are close to 1 which further confirms convergence.


```{r,warning=FALSE,message=FALSE}
# Extract RR 
RR=jags.mod.fit$BUGSoutput$mean$RR

# Extra bit
scot$RR=RR
scot$diff = smr - RR

# Compare mean of RR to SMR
mean(RR)
mean(smr)

# Create a data frame to plot the difference between RR and smr
df= data_frame(smr=smr,RR=RR)
ggplot(df,) + geom_density(aes(x=smr,color="blue")) +
  geom_density(aes(x=RR,color="red"))


# Map RR and smr
ScotlandMap(RR,figtitle="Scotland RR")
ScotlandMap(smr,figtitle="Scotland smr")



ggplot(scot,) + geom_density(aes(x=diff,color="red"))


```

The posterior means for the RR and the smr are very similar with the RR values generally lower indicating that the SMR overestimates the actual RR.

```{r,warning=FALSE,message=FALSE}

# Specify model in JAGS
jags.mod2 <- function(){
  # Likelihood
  for (i in 1:n) {
    Obs[i] ~ dpois(mu[i])
    log(mu[i]) <- log(Exp[i]) + beta0 + theta[i]
    theta[i] ~ dnorm(0, tau)
    RR[i] <- exp(beta0 + theta[i])
  }
  
  # Prior for beta0
  beta0 ~ dnorm(0, 1/1000)
  
  
  
  tau ~ dgamma(1, 0.05)
  
  
  # Add node of probability that RR > 1
  RRpred = ifelse(RR > 1, 1 , 0)
  
  
}


# Parameters we want to monitor
jags.param2 <- c('RR',"RRpred")



# Fit the JAGS model
jags.mod.fit2 = jags(data = jags.data , inits = jags.inits,
                    parameters.to.save = jags.param2, 
                    n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 2500,n.thin=1, 
                    model.file = jags.mod2,
                    DIC=TRUE)


# Extract th probabilities
RRpred = jags.mod.fit2$BUGSoutput$mean$RRpred

# Map the probs
ScotlandMap(RRpred,figtitle="RRpred")

# Extract probs where 95% certainty

a95 =  ifelse(RRpred > 0.95,1,0)
ScotlandMap(a95,figtitle="Areas>95% certain contains excess cancer")


# Find areas with RR_prob >= 0.95
scot$RRpred= RRpred
excess_risk_areas <- scot$Name[scot$RRpred >= 0.95]

# Print the names of these areas
print(excess_risk_areas)


```



```{r,warning=FALSE,message=FALSE}


# Specify model in JAGS
jags.mod3 <- function(){
  # Likelihood
  for (i in 1:n) {
    Obs[i] ~ dpois(mu[i])
    log(mu[i]) <- log(Exp[i]) + beta0 + theta[i]
    theta[i] ~ dnorm(0, tau)
    RR[i] <- exp(beta0 + theta[i])
  }
  

  
  # Prior for intercept
  beta0 ~ dnorm(0, 0.1)
  
  # Prior for precision parameter
  tau ~ dgamma(2, 0.01)
}



# Fit the JAGS model
jags.mod.fit3 = jags(data = jags.data , inits = jags.inits,
                    parameters.to.save = jags.param, 
                    n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 2500,n.thin=1, 
                    model.file = jags.mod3, 
                    DIC=TRUE)


RR3 = jags.mod.fit3$BUGSoutput$mean$RR

# Map both RRs and observe differences 

ScotlandMap(RR3,figtitle="Scotland RR3")
ScotlandMap(RR,figtitle="Scotland RR")

summary(jags.mcmc3)


```


The priors chosen for this model are vague priors. With Beta0 having normal prior with (0,10) and tau having a gamma prior (2,0.01). For Beta0 a 0 mean shows no prior preference for the intercept value, the variance of 10 still makes it a vague prior but more informative than the prior with variance 1000. This will make the values more concentrated towards the mean but still allow for estimates to be made.

For the tau the gamma(2,0.01) has a mean of 200 which is much higher than the previous prior but still vague. The shape parameter is set as 2 instead of 1 to give more weight to lower vlaues of the distribution and allow for uncertainity.



```{r,warning=FALSE,message=FALSE}

# Jags model

jags.mod4 <- function(){ for(i in 1:12){
  r[i] ~ dbin(theta[i],n[i])
  logit(theta[i]) <- logit.theta[i]
  logit.theta[i] ~ dnorm(mu, tau)
}
  mu ~ dnorm(0,1.0E-3)
  tau <- 1/sigma^2
  sigma ~ dunif(0,100)
}


# Specify initial values
inits4 <-  list('mu'= 0,"sigma"=0,"logit.theta"=rep(0,12))
inits4 <- list('mu'= 1,"sigma"=1,"logit.theta"=rep(1,12))
jags.inits4 <- list(inits4, inits4)


# Parameters we want to monitor
jags.param4 <- c('theta' ,'mu',"sigma")


# Fit the JAGS model
jags.mod.fit4 = jags(data=as.list(surg) , inits = jags.inits4,
                    parameters.to.save = jags.param4, 
                    n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 5000,n.thin=1, 
                    model.file = jags.mod4,
                    DIC=TRUE)



print(jags.mod.fit4)
jags.mcmc4=as.mcmc(jags.mod.fit4)
gelman.diag(jags.mcmc4,multivariate=F)
gelman.plot(jags.mcmc4)
MCMCtrace(jags.mcmc4)
summary(jags.mcmc4)

```

Looking at the traceplots all paramters have converged since plots show good mixing, this is futher confirmed by the gelman rubin diagnostic which shows all values very close to 1.



```{r,warning=FALSE,message=FALSE}


# Jags model

jags.mod5 <- function(){ for(i in 1:12){
  r[i] ~ dbin(theta[i],n[i])
  logit(theta[i]) <- logit.theta[i]
  logit.theta[i] ~ dnorm(mu, tau)
  rpred[i] ~ dbin(theta[i],n[i])
}
  mu ~ dnorm(0,1.0E-3)
  tau <- 1/sigma^2
  sigma ~ dunif(0,100)
  midp = ifelse(rpred > r,1,0) + 0.5 * ifelse(rpred==r,1,0)
}



# Parameters we want to monitor
jags.param5 <- c("rpred","midp","theta")

# Specify initial values
inits5 <-  list('mu'= 0,"sigma"=0,"logit.theta"=rep(0,12),"r.pred"=rep(30,12))
inits5 <- list('mu'= 1,"sigma"=1,"logit.theta"=rep(1,12),"r.pred"=rep(32,12))
jags.inits5 <- list(inits5, inits5)


# Fit the JAGS model
jags.mod.fit5 = jags(data=as.list(surg) , inits = jags.inits5,
                    parameters.to.save = jags.param5, n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 5000,n.thin=1, model.file = jags.mod5,
                    DIC=TRUE)

print(jags.mod.fit5)

# Convert to mcmc and get summaries
jags.mcmc5 = as.mcmc(jags.mod.fit5)
summary(jags.mcmc5)


# Observe traceplots
MCMCtrace(jags.mcmc5,type="density",ind = T)


# Extract R and rpred 
rpred = jags.mod.fit5$BUGSoutput$mean$rpred
r=surg$r

# Create a data frame to plot the difference between r and rpred
df=data_frame(r=r,rpred=rpred)
ggplot(df,)+geom_density(aes(x=r,color="blue"))+
  geom_density(aes(x=rpred,color="red"))


# Observe and revise midp values
jags.mod.fit5$BUGSoutput$mean$midp




```




```{r,warning=FALSE,message=FALSE}
# Jags model

jags.mod6 <- function(){ for(i in 1:12){
  r[i] ~ dbin(theta[i],n[i])
  theta[i] ~ dunif(0, 1)
}
  mu ~ dnorm(0,1.0E-3)
  tau <- 1/sigma^2
  sigma ~ dunif(0,100)
}

# Specify initial values
inits6 <-  list('mu'= 0,"sigma"=0)
inits6 <- list('mu'= 1,"sigma"=1)
jags.inits6 <- list(inits6, inits6)

# Parameters we want to monitor
jags.param6 <- c("theta")


# Fit the JAGS model
jags.mod.fit6 = jags(data=as.list(surg) , inits = jags.inits6,
                    parameters.to.save = jags.param6, n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 5000,n.thin=1, model.file = jags.mod6, 
                    DIC=TRUE)



print(jags.mod.fit6)
jags.mcmc6=as.mcmc(jags.mod.fit6)
summary(jags.mcmc6)


# Extract theta from fixed and random effects models

rndm=jags.mod.fit4$BUGSoutput$mean$theta
fixed=jags.mod.fit6$BUGSoutput$mean$theta
fixed
rndm

df= data_frame(rndm=rndm,fixed=fixed)


# combine posterior samples into one data frame
theta_all <- rbind(
  data.frame(theta = rndm, model = "random effects"),
  data.frame(theta = fixed, model = "fixed effects")
)

# plot boxplots of mortality rates for each hospital under each model
ggplot(theta_all, aes(x = factor(model), y = theta, 
                      fill = factor(model))) +
  geom_boxplot() +
  xlab("Model") +
  ylab("Mortality rate") +
  ggtitle("Comparison of posterior distributions for hospital mortality rates") +
  theme_bw()

boxplot(rndm)


## The 11th hospital looks like an outlier in both models

```



```{r,warning=FALSE,message=FALSE}

# Jags model

jags.mod7 <- function(){ for(i in 1:12){
  r[i] ~ dbin(theta[i],n[i])
  theta[i] ~ dunif(0, 1)
  rpred[i] ~ dbin(theta[i],n[i])
}
  mu ~ dnorm(0,1.0E-3)
  tau <- 1/sigma^2
  sigma ~ dunif(0,100)
  midp = ifelse(rpred > r,1,0) + 0.5 * ifelse(rpred==r,1,0)
}



# Fit the JAGS model
jags.mod.fit7 = jags(data=as.list(surg) , inits = jags.inits4,
                    parameters.to.save = jags.param5, 
                    n.chains = 2,
                    n.iter = 10000,
                    n.burnin = 5000,n.thin=1, 
                    model.file = jags.mod7, 
                    DIC=TRUE)



print(jags.mod.fit7)
jags.mcmc7=as.mcmc(jags.mod.fit7)
gelman.diag(jags.mcmc7,multivariate=F)
summary(jags.mcmc7)


midprndm=jags.mod.fit5$BUGSoutput$mean$midp
midpfxd=jags.mod.fit7$BUGSoutput$mean$midp

midpfxd  
midprndm 

par(mfrow = c(1, 2))
boxplot(midprndm  , main = "Random Effects", ylab = "Mid P-value")
boxplot(midpfxd , main = "Fixed Effects", ylab = "Mid P-value")



```




